Fused regression for network inference in multiple species

### 2015-08-30 paper1.tex
ahh
hi

There exist several tools for identifying orthologous genes using sequence similarity. Although sequence similarity is a useful predictor of functional similarity, we recognize that some genes have taken on different functions and therefore may have new regulatory interactions. When this occurs, the benefits in overall performance from our L2 fusion approach come at the cost of impaired prediction of regulators for these genes. The constraints we impose favor weight configurations in which fused interaction weights are similar to one another. Our fused L2 method minimizes a cost function that strives to simultaneously fit expression data in each species and produce networks that are consistent with these evolutionary constraints. When a gene's regulators have changed through evolution, there may be no way of achieving both these goals. Similarly, if a gene has different regulators during different processes, the expression data may conflict with the fusion constraints. In these cases, the network inference procedure would be improved by relaxing or ignoring these evolutionary constraints. However, we don't know a priori which constraints to enforce and which to relax.

We approach this problem by introducing a saturating penalty function based on work in statistics to develop unbiased regularization penalties for regression. The resulting cost function is non-convex and difficult to optimize and fares less well in the underdetermined case; however, we can approximate its solution and obtain deeper insights into functional similarity than is available through orthology or the comparison of separately learned networks. When the quantity of data is sufficient to produce high quality networks, our adaptive fusion approach improves performance of network inference; this is achieved in part through the identification of presumptive orthologs with different functions. In simulated data, when we introduced faulty orthology information, our method not only correctly identified which evolutionary restraints to enforce and which to relax, but also outperformed the control case where no incorrect orthologs were used.

>>>>

There exist several tools for predicting orthology between genes using sequence similarity. Although sequence similarity is a useful predictor of functional similarity, we recognize that some genes have taken on different functions and therefore may have new regulatory interactions. When this occurs, the benefits in overall performance from our L2 fusion approach come at the cost of impaired prediction of regulators for these genes. The constraints we impose favor weight configurations in which fused interaction weights are similar to one another. Our fused L2 method minimizes a cost function that strives to simultaneously fit expression data in each species and produce networks that are consistent with these evolutionary constraints. When a gene's regulators have changed through evolution, there may be no way of achieving both these goals. Similarly, if a gene has different regulators during different processes, the expression data may conflict with the fusion constraints. In these cases, the network inference procedure would be improved by relaxing or ignoring these evolutionary constraints. However, we don't know a priori which constraints to enforce and which to relax.

We approach this problem by introducing a saturating penalty function based on work in statistics to develop unbiased regularization penalties for regression. The main difference between the L2 fusion approach and this adaptive fusion approach occurs when the difference between presumed analogous interactions is large despite the fusion penalty: we assume that large differences represent evidence that functions have diverged and represent this, in adaptive fusion, with a relaxation of the fusion penalty. The resulting cost function is non-convex and difficult to optimize; however, we can approximate its solution and obtain deeper insights into functional similarity than is available through orthology or the comparison of separately learned networks.
